% the acknowledgments section
% advisors Les and Ryan

%This thesis would not exist without the support and
%encouragement of many wonderful people. 
%Foremost among these are my advisors, Leslie Valiant
%and Ryan Adams.
\vspace{-1em}
I am extraordinarily fortunate to have had two fantastic advisors, Leslie Valiant and Ryan Adams.
%First, I would like to thank my advisors, Leslie Valiant and Ryan Adams.
Les has been an amazing source of
wisdom, perspective, and inspiration. I am thankful for
his patience and his guidance in sifting through
the excitement of the moment and identifying lasting
questions. 
The first class I attended at Harvard was Ryan's first lecture as a professor,
and that fortuitous meeting kicked off five of the most
rewarding years of my life.
%I continue to marvel at the energy and creativity Ryan brings to science.
Of all that he has shared with me~---~a wealth of technical knowledge,
an ability to articulate ideas, and an appreciation for
clear figures and neat handwriting~---~I am most grateful for the joy he has helped me find in research.

% Jonathan Pillow, Sam Gershman, Haim Sompolinsky
I also wish to thank Jonathan Pillow, Sam Gershman, and Haim
Sompolinsky, who have been invaluable sources of mentorship and
advice.  Their generous sharing of time and ideas has had a profound
impact on my perspective of computational neuroscience.

%Jonathan has introduced me to a fantastic community of
%computational neuroscientists at Cosyne and Woods Hole, and has been
%an awesome collaborator and sounding board for many of the projects in
%this thesis.  Sam has taken me under his wing over the past year and
%taught me to ask the hard questions and think about the broader
%applications and implications of my work.  In Haim's course on
%theoretical neuroscience, which was by far the best course I've ever
%taken, I learned to be rigorous and unafraid of complex
%systems. Moreover, I am thankful for his generosity in meeting with me
%throughout my Ph.D. and for welcoming me into the computational
%community he has fostered in the Center for Brain Science.


% HIPS Group - David Duvenaud, Miguel Hernandez Lobato, Dougal Maclaurin, Andy Miller, Oren Rippel, Yakir Reshef, Jasper Snoek, Alex Wiltschko
% Mike, Elaine, Robert, Alex, Eyal, Jonathan, Finale, Shamim
I cannot imagine a more fun and intellectually stimulating community
than the Harvard Intelligent Probabilistic Systems group.  I am
grateful to David Duvenaud, Miguel Hernandez-Lobato, Dougal Maclaurin,
Andy Miller, Oren Rippel, Yakir Reshef, and Jasper Snoek for all the
``unsupervised learning'' I've benefitted from in MD209.
% Matt
I'd particularly like to thank Matt Johnson, who has literally taught
me everything I know about many subjects.  His enthusiasm and
encouragement have made me a better researcher and a stronger person.

% Aaron and Chris
I am indebted to Chris Stock and Aaron Tucker for working with me
during their undergraduate years. Whatever I may have taught them
about computational neuroscience surely pales
in comparison to what they taught me about advising. Their
hard work has shaped many pages of this thesis. 

% Friends in theoretical CS: Varun, Thomas, Mark, Jon, Justin, James
% Housemates at Sashton 
I thank my friends in the theory group --- Varun, Thomas,
Mark, Jon, and Justin --- for pushing me to think abstractly, and I
thank my housemates in Cambridge --- Ryan, Bill, James, Chris,
Eddie, and Steve --- for keeping me grounded.
My graduate years have been blessed by many other wonderful friends, 
and I thank them for all of the fond memories.

% Funding
%I have benefitted from the generous support of the National
%Defense Science and Engineering Graduate (NDSEG) Fellowship, the
%NSF Center for Brains, Minds, and Machines (CBMM), and the Siebel
%Foundation. 

% Family
My parents, Richard and Linnea, have always encouraged me to pursue
this dream, and I am forever grateful for all the opportunities they
have given me. I thank my brothers, Stephen, Matthew, and Randy, for
never letting me miss a beat.  To Stephen, who once warned me that if
I delayed on graduate school I might not finish before the age of
thirty, I say: five months to spare!

% Anne 
Above all, my thanks go to my wife, Anne, who has been my partner
throughout this great adventure. This thesis is a testament to
her unbounded love and support. 
